---
title: "BSM"
output: html_document
date: "2023-11-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Defining a BSM model with rjd3sts

Thie package allows several equivalent definitions of a basic strctural model.
We present below some of them.

To compare the results (more precisely the likelihood) of the different approaches,
it is important to compute the marginal likelihood.

#### Standard definition, noise in the state

```{r}
library(rjd3sts)
library(rjd3toolkit)

s<-log(retail$BookStores)

# create the model
bsm<-model()
# create the components and add them to the model
add(bsm, locallineartrend("ll"))
add(bsm, seasonal("s", 12, type="HarrisonStevens"))
add(bsm, noise("n"))
rslt<-estimate(bsm, log(s), marginal=T)

print(result(rslt, "likelihood.ll"))
print(result(rslt, "parameters"))


```
#### Standard definition, noise in the measurement

```{r}

# create the model
bsm<-model()
# create the components and add them to the model
add(bsm, locallineartrend("ll"))
add(bsm, seasonal("s", 12, type="HarrisonStevens"))
  # create the equation (fix the variance to 1)
eq<-equation("eq", 1,T)
add_equation(eq, "ll")
add_equation(eq, "s")
add(bsm, eq)
rslt<-estimate(bsm, log(s), marginal=T)

print(result(rslt, "likelihood.ll"))
print(result(rslt, "parameters"))


```
#### components with fixed variances, aggregated with diffuse weights (noise in the state)

```{r}

# create the model
bsm<-model()
  # create the components, with fixed variances, and add them to the model
add(bsm, locallineartrend("ll", 
                             levelVariance = 1, fixedLevelVariance = TRUE) )
add(bsm, seasonal("s", 12, type="HarrisonStevens", 
                     variance = 1, fixed = TRUE))
add(bsm, noise("n", 1, fixed=TRUE))
  # create the equation (fix the variance to 1)
eq<-equation("eq", 0, TRUE)
add_equation(eq, "ll", .01, FALSE)
add_equation(eq, "s", .01, FALSE)
add_equation(eq, "n")
add(bsm, eq)
rslt<-estimate(bsm, log(s), marginal=TRUE)

print(result(rslt, "likelihood.ll"))
p<-result(rslt, "parameters")
print(p)

cat ('to be noted:')

cat ('p[5]*p[5]=l-var=', p[5]*p[5])
cat ('p[5]*p[5]*p[2]=slope-var=', p[5]*p[5]*p[2])
cat ('p[6]*p[6]=s-var=', p[6]*p[6])

```

#### bsm with long term trend and cycle

```{r}

# create the model
bsm<-model()
  # create the components and add them to the model
add(bsm, locallevel("l", initial = 0) )
add(bsm, locallineartrend("lt", levelVariance = 0, 
                             fixedLevelVariance = TRUE) )
add(bsm, seasonal("s", 12, type="HarrisonStevens"))
add(bsm, noise("n", 1, fixed=TRUE))
  # create the equation (fix the variance to 1)
rslt<-estimate(bsm, log(s), marginal=TRUE)

print(result(rslt, "likelihood.ll"))
print(result(rslt, "parameters"))

ss<-smoothed_states(rslt)
plot(ss[,1]+ss[,2], type='l', col='blue', ylab='trends')
lines(ss[, 2], col='red')

```
